### 2018-06-11 express2018/formal.tex
Stage work on part2

%%%% -*- Mode: LaTeX -*-
%%
%% This is the draft of the 2nd part of EXPRESS/SOS 2018 paper, co-authored by
%% Prof. Davide Sangiorgi and Chun Tian.

@@

In the second half of this paper, we present a formalization of CCS
including the new concepts and theorems proposed in the first half of
this paper. The main purpose is to convince the readers that, there's no flaw in the informal proofs. The
formalization work is based on a formalization of CCS in HOL theorem
prover (HOL4), described in second author's MSc thesis
\cite{Tian:2017wrba} and now lives in the official examples folder of HOL4 source
code\footnote{\url{https://github.com/HOL-Theorem-Prover/HOL}}.
The work is essentially a porting and extension of the early CCS
formalization by Monica Nesi \cite{Nesi} in HOL88.

@@

But such a simplification doesn't hurt the rigorousness of our entire
work, as once the single-equation version of the theorem is
proven, the general case is just a routine
adaptation for paper proofs. \footnote{However, it's impossible to directly derive the
  multi-variable versions of unique solution theorems from their
  single-variable versions.} \footnote{Also, I don't know if there
  exists a higher-level general
  duality theorem such that some kind of results on single-variable CCS
  equations must also hold for multi-variable equation groups. --Chun.}
It's not easy to formalize multi-variable CCS equations with
introducing a large number of definitions and intermediate
results. It's easy, however, to directly use the semantics context to
represent single-variable equations without the need of formalizing
the concept of ``equation'' at all.

There're two major limitations in this work. One is the class of CCS
processes, i.e. we're limited with Finitary CCS which has no infinite
sums or parallel compositions. The other limitation is, even for single-variable equation, the
variable doesn't appear inside any constant.

\subsection{CCS, Transitions and Bisimilarities}

In our formalization of CCS, the set of transition actions are not
defined by the union of all visiable actions (names and conames) and the invisible
action $\tau$. Instead, we define only visible names as a simple
algebraic datatype ``\HOLinline{\ensuremath{\beta} \HOLTyOp{Label}}`` in which the type parameter
$\beta$ can be any other type (thus we don't know anything about the
cardinality of the set of all visiable actions), and then use HOL's
\texttt{optionTheory} to further extend the type by one more
element, $\tau$. Having a dedicated type for all visiable actions is quite
convenient and make some statements shorter. (e.g. the restriction
operator can only take a set of visible actions)

The type of CCS process is defined by an algebraic datatype in HOL,
i.e. an inductive defined finite structure constructed by unary and
binary operators.\footnote{It seems possible to switch to co-algebraic
datatypes for the support of potential infinite sums. Indeed, we never
need to induct on CCS processes in any proof so far.} 
It's not easy to directly extend this datatype with new operators,
without hurting the existing proved results.\footnote{This is usually
  called \emph{deep embedding} when implementing a new logic in theorem
  provers. In case of CCS, \emph{shallow embedding} seems impossible.}
Relabeling operator is not necessary from the theoretical view, as
it's possible to use \emph{syntactic
  substitution} instead\footnote{c.f. Section 4.1.2. (page 168) of Robert Gorrieri's CCS book.}
However, it's quite hard to formalize syntactic substitution for CCS,
as it requires to prove many results on the free and bound names
first.

Without constants (or recursion operator), a CCS process cannot
represent infinite-state LTS and many deep results in CCS cease to
exist then. Therefore any serious CCS formalization
must contain the support of constants (or recursion operator. Nowadays
most CCS textbooks and papers use constants, but from the formal view
it's better (or only possible) to use the original form: the recursion
operator, because any CCS process must be presented as single, finite term in
theorem prover. In Milner's book, this CCS operator is called \texttt{fix},
as it essentially returns the fixpoint of a CCS equation. We use the
symbol \texttt{rec} instead and limit ourself to single-variable case,
as we want to extend to multi-variable recursion operator in the
future with the original name. So instead of saying a CCS process $A$
with constant definitions
$A \overset{\mathrm{def}}{=} \tau.A + b.B$ and $B
\overset{\mathrm{def}}{=} b.B$, we use $\mathbf{rec}\;A.\, (\tau.A +
b.(\mathbf{rec}\;B.\,(b.B)))$ instead.

The extra beauty of above recursion opeator is, if we imagine the
\texttt{rec} symbol as $\lambda$ in $\lambda$-expressions, i.e. $\lambda A.\, (\tau.A +
b.(\lambda B.\,(b.B)))$, then we can see both $A$ and $B$ are bound
variables in this expression. On the other side, it's easy to write
an extra constant $C$ inside previous process without a surrounding $\lambda C$, and it must
be a \emph{free variable} of that $\lambda$-expressions. If we go back to the constant-based CCS
syntax, this would mean a malformed CCS process because not all the
constants used in it are defined. Our point here is:
\begin{enumerate}
\item Under the recursion-based CCS syntax, all possible CCS terms
  constructed from the algebraic datatype are well-formed, and those
  free variables can be seen and used\footnote{However, in our
    formalization of unique solution of equations theorems, we focus
    on single-variable case and directly used $\lambda$-expression
    (typed: $CCS\rightarrow CCS$) as equation. The current discussion
    is mainly for the support of multi-variable equations in the future.} as equation variables.
\item All proved theorems about CCS naturally take CCS process with
  extra free variables, as they by themselves have no transition, just
  like $\mathrm{0}$. (this is a natural consequence derived from SOS rules)
\end{enumerate}

The transition of CCS processes is defined by an inductive
relation, literally the same as SOS rules appearing in CCS
textbooks.\footnote{There's an interesting finding here: we never need to use the induction
theorem for transition relation at all, after having proved almost all
notable results for CCS. In another word, the whole CCS formal theory
builds normally, even if the transition relation
is defined co-inductively (by the same set of SOS rules). A potential
need is the prove the set of ending process of transitions from any
process doesn't use any new labels, but so far this result is not
formalized yet. The question
is: do the largest fixpoint and smallest fixpoint \emph{coincide} for
a relation defined by SOS rules? (if not, what else do we have in the
co-inductive relation?)} The only notable SOS rule which is different
with their usual version in textbooks, is the recursion rule:
\begin{alltt}
\HOLTokenTurnstile{} \HOLConst{CCS_Subst} \HOLFreeVar{E} (\HOLConst{rec} \HOLFreeVar{X} \HOLFreeVar{E}) \HOLFreeVar{X} \HOLTokenTransBegin\HOLFreeVar{u}\HOLTokenTransEnd \HOLFreeVar{E\sb{\mathrm{1}}} \HOLSymConst{\HOLTokenImp{}} \HOLConst{rec} \HOLFreeVar{X} \HOLFreeVar{E} \HOLTokenTransBegin\HOLFreeVar{u}\HOLTokenTransEnd \HOLFreeVar{E\sb{\mathrm{1}}}
\end{alltt}
which says, if we substitute all appearences of constant $X$ in $E$ to
$(\mathbf{rec}\; X E)$ and the resulting process has a transition to $E_1$
with action $u$, then $(\mathbf{rec}\; X E)$ has the same
transition.

\subsection{Coarsest (pre)congruence contained in $\approx$ (or $\succeq_{\mathrm{bis}}$)}

The ``coarsest congruence contained in weak bisimilarity ($\approx$)''
theorem in CCS is somehow special, as its current known proofs either
rely on quite restricted conditions, or have an extremely complicated proof
(c.f. van Glabbek's paper) in
which ordinal theory is required.  Actually even the relationship
between its name and statement is not well explained in many CCS
textbooks. But van Glabbek's paper has given the so far clearest
explainion, here we briefly repeat his arguments:

As we know weak bisimilarity ($\approx$) is not (real) congruence, as
it doesn't satisfy subsitutivity on direct sums (but if the CCS syntax
is non-standard, i.e. has only prefixed sums, $\approx$ is indeed a
congruence). The purpose is to find a coarsest congruence contained in
weak bisimilarity. (``coarsest'' means, any other congruence finer than it must be contained in it)
There're two ways to build a congruence from weak bisimilarity, one
way is the standard definition for observational congruence (rooted
weak bisimilarity) $\approx^c$ in CCS textbooks, but even it's proven to be a
congruence we don't know if it's coarsest one.  The other way is to
build a (pre)congruence closure (Def ??) directly upon
the original weak bisimilarity relation, we call the resulting
relation ``Weak bisimilarity congruence'' ($[\approx]$):
\begin{alltt}
\HOLConst{WEAK_CONGR} \HOLSymConst{=} \HOLConst{CC} \HOLConst{WEAK_EQUIV}
\HOLConst{CC} \HOLFreeVar{R} \HOLSymConst{=} (\HOLTokenLambda{}\HOLBoundVar{g} \HOLBoundVar{h}. \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{c}. \HOLConst{CONTEXT} \HOLBoundVar{c} \HOLSymConst{\HOLTokenImp{}} \HOLFreeVar{R} (\HOLBoundVar{c} \HOLBoundVar{g}) (\HOLBoundVar{c} \HOLBoundVar{h}))
\end{alltt}
It can be shown that any such (pre)congruence closure is automatically coarset.

Now it remains to prove that, the congrunce relation built by above
two quite different approaches actually coincide. To achive this goal,
we first noticed that, all other operators beside sums used in
semantic context doesn't matter, because they're already substituible
for weak bisimilarity. The only important operator is the sum
operator. To focus on this important operator, we can temporily
introduce another concept called \emph{sum equivalence}:
\begin{alltt}
\HOLConst{SUM_EQUIV} \HOLSymConst{=} (\HOLTokenLambda{}\HOLBoundVar{p} \HOLBoundVar{q}. \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{r}. \HOLBoundVar{p} \HOLSymConst{+} \HOLBoundVar{r} \HOLSymConst{\HOLTokenWeakEQ} \HOLBoundVar{q} \HOLSymConst{+} \HOLBoundVar{r})
\end{alltt}
It can be shown that, weak bisimilarity congruence implies this sum
equivalence. Thus if we can further prove that the sum
equivalence imples observational congrence, all the three relations
must coincide together, as shown in the following diagram:
\begin{displaymath}
\xymatrix{
{\textrm{Weak bisimilarity } (\approx)} & {\textrm{Sum
    equivalence } (\approx^+)} \ar@/^3ex/[ldd]^{\subseteq ?}\\
{\textrm{Weak bisim. congruence } ([\approx])}
\ar[u]^{\subseteq} \ar[ru]^{\subseteq} \\
{\textrm{Rooted bisimilarity } (\approx^c)} \ar[u]^{\subseteq}
}
\end{displaymath}
In another words, observational congruence is the coarsest congruence
contained in weak bisimilary if and only if
\begin{equation}
\forall p\; q.\; p \approx^c\! q \Longleftrightarrow \forall r.\; p+r \approx
q+r
\end{equation}

The classical assumption for proving above theorem requires that $p$
and $q$ do not use up all available visible actions,
i.e. $\mathrm{fn}(p) \cup \mathrm{fn}(q) \neq \mathscr{L}$. But it's
not easy to formalize and use such an assumption without a detailed
treatment on free and bound names (visible actions) of CCS
processes.\footnote{There're totally four such concepts: 1) free names
are all visible actions appearing in a CCS term without surrounding
$\nu$ (restriction) operator on the same action; 2) bound names are
the set of all actions ever used by $\nu$ (restriction) operator; 3)
free variables (or equation variables) are those variables without a
definition given by recursion
operator; 4) bound variables (process constants) are variables with
definitions given by recursion operator. All CCS results using these
concepts are not touched so far, although these four concepts are
successfully defined using HOL's set and list theories.} However, by
analyzing the proof steps, we found that, what's really required is to
not use up all available labels in those weak transitions directly
lead from $p$ and $q$. In another words, even they have used all
available labels, as long as their first weak transitions didn't, the
whole proof can still be finished.\footnote{Further more, $p$ and $q$
  can be considered separately: the proof can be finished as long as
  \emph{each} of them didn't use up all labels on first weak
  transition, while the union of these labels are all labels.}
We have formalized this property of
CCS process and call it ``free action'' property:
\begin{alltt}
\HOLTokenTurnstile{} \HOLConst{free_action} \HOLFreeVar{p} \HOLSymConst{\HOLTokenEquiv{}} \HOLSymConst{\HOLTokenExists{}}\HOLBoundVar{a}. \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{p\sp{\prime}}. \HOLSymConst{\HOLTokenNeg{}}(\HOLFreeVar{p} \HOLTokenWeakTransBegin\HOLConst{label} \HOLBoundVar{a}\HOLTokenWeakTransEnd \HOLBoundVar{p\sp{\prime}})
\end{alltt}
With this property, the classical form of this theorem that we have
formally proved is:
\begin{alltt}
\HOLTokenTurnstile{} \HOLConst{free_action} \HOLFreeVar{p} \HOLSymConst{\HOLTokenConj{}} \HOLConst{free_action} \HOLFreeVar{q} \HOLSymConst{\HOLTokenImp{}}
   (\HOLFreeVar{p} \HOLSymConst{\HOLTokenObsCongr} \HOLFreeVar{q} \HOLSymConst{\HOLTokenEquiv{}} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{r}. \HOLFreeVar{p} \HOLSymConst{+} \HOLBoundVar{r} \HOLSymConst{\HOLTokenWeakEQ} \HOLFreeVar{q} \HOLSymConst{+} \HOLBoundVar{r})
\end{alltt}

If we drop this classical assumption, then the proof becomes much
harder, as given a van Glabbek's paper. The most important
intermediate result we have proved here, is the following lemma:
\begin{alltt}
\HOLTokenTurnstile{} (\HOLSymConst{\HOLTokenExists{}}\HOLBoundVar{k}.
        \HOLConst{STABLE} \HOLBoundVar{k} \HOLSymConst{\HOLTokenConj{}} (\HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{p\sp{\prime}} \HOLBoundVar{u}. \HOLFreeVar{p} \HOLTokenWeakTransBegin\HOLBoundVar{u}\HOLTokenWeakTransEnd \HOLBoundVar{p\sp{\prime}} \HOLSymConst{\HOLTokenImp{}} \HOLSymConst{\HOLTokenNeg{}}(\HOLBoundVar{p\sp{\prime}} \HOLSymConst{\HOLTokenWeakEQ} \HOLBoundVar{k})) \HOLSymConst{\HOLTokenConj{}}
        \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{q\sp{\prime}} \HOLBoundVar{u}. \HOLFreeVar{q} \HOLTokenWeakTransBegin\HOLBoundVar{u}\HOLTokenWeakTransEnd \HOLBoundVar{q\sp{\prime}} \HOLSymConst{\HOLTokenImp{}} \HOLSymConst{\HOLTokenNeg{}}(\HOLBoundVar{q\sp{\prime}} \HOLSymConst{\HOLTokenWeakEQ} \HOLBoundVar{k})) \HOLSymConst{\HOLTokenImp{}}
   (\HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{r}. \HOLFreeVar{p} \HOLSymConst{+} \HOLBoundVar{r} \HOLSymConst{\HOLTokenWeakEQ} \HOLFreeVar{q} \HOLSymConst{+} \HOLBoundVar{r}) \HOLSymConst{\HOLTokenImp{}}
   \HOLFreeVar{p} \HOLSymConst{\HOLTokenObsCongr} \HOLFreeVar{q}
\end{alltt}
It roughly says, for any two processes $p$ and $q$, if we can find
another stable (no $\tau$-transitions) process $k$ which is not weak bisimilar to any transition of
$p$ and $q$, then the hard part of our main theorem is proved. In
practice, once two processes were given, it's not hard to find
such a process, but the arbitrariness of $p$ and $q$ made this result
extremely hard to prove. The method given by van Glabbek requires a
construction of arbitrarily non-bisimilar processes called ``Klop
processes'':
\begin{definition}{(Klop processes)}
For each ordinal $\lambda$, and an arbitrary chosen non-$\tau$ action $a$,
define a CCS process $k_\lambda$ as follows:
\begin{enumerate}
\item $k_0 = 0$,
\item $k_{\lambda+1} = k_\lambda + a.k_\lambda$ and
\item for $\lambda$ a limit ordinal, $k_\lambda = \sum_{\mu < \lambda}
  k_\mu$, meaning that $k_\lambda$ is constructed from all graphs
  $k_\mu$ for $\mu < \lambda$ by identifying their root.
\end{enumerate}
\end{definition}
It's not hard to prove that, all $k_i$ are non-bisimilar (not only
weakly but also strongly). The idea is, for any two processes, the union of sets of their all
transitions cannot be arbitrarily large: it has to be limited by an
ordinal number. But above contruction can be arbitrarily large, so
there always exists a Klop process which can be used to satisfy above
lemma (and finish the proof).

However, this goes beyond HOL's expressivity to define above process, mostly
because there's no way to express infinite ``sums'' in CCS
datatype.\footnote{There're actually several ways to modify the
  datatype to support infinite sums, but none of these infinity can be
arbitrary. The full version of ordinal theory cannot be expressed in HOL.}
What we can do is to eliminate the last branch with limiting
ordinals, and define Klop processes only on finite cases:
\begin{alltt}
\HOLConst{KLOP} \HOLFreeVar{a} \HOLNumLit{0} \HOLSymConst{=} \HOLConst{nil}
\HOLConst{KLOP} \HOLFreeVar{a} (\HOLConst{SUC} \HOLFreeVar{n}) \HOLSymConst{=} \HOLConst{KLOP} \HOLFreeVar{a} \HOLFreeVar{n} \HOLSymConst{+} \HOLConst{label} \HOLFreeVar{a}\HOLSymConst{..}\HOLConst{KLOP} \HOLFreeVar{a} \HOLFreeVar{n}\hfill[KLOP_def]
\end{alltt}
But this means we much assume the two processes $p$ and $q$ are
finite-state, i.e. their corresponding LTS graphs have only finite
states. Choosing an element from a countable infinite set of processes
to make sure it's not bisimiar with a given set of processes, this is
essentially a pure set-theoretic problem, as formalized and proved in
the following lemma:
\begin{alltt}
\HOLTokenTurnstile{} \HOLConst{equivalence} \HOLFreeVar{R} \HOLSymConst{\HOLTokenImp{}}
   \HOLConst{FINITE} \HOLFreeVar{A} \HOLSymConst{\HOLTokenConj{}} \HOLConst{INFINITE} \HOLFreeVar{B} \HOLSymConst{\HOLTokenConj{}}
   (\HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{x} \HOLBoundVar{y}. \HOLBoundVar{x} \HOLSymConst{\HOLTokenIn{}} \HOLFreeVar{B} \HOLSymConst{\HOLTokenConj{}} \HOLBoundVar{y} \HOLSymConst{\HOLTokenIn{}} \HOLFreeVar{B} \HOLSymConst{\HOLTokenConj{}} \HOLBoundVar{x} \HOLSymConst{\HOLTokenNotEqual{}} \HOLBoundVar{y} \HOLSymConst{\HOLTokenImp{}} \HOLSymConst{\HOLTokenNeg{}}\HOLFreeVar{R} \HOLBoundVar{x} \HOLBoundVar{y}) \HOLSymConst{\HOLTokenImp{}}
   \HOLSymConst{\HOLTokenExists{}}\HOLBoundVar{k}. \HOLBoundVar{k} \HOLSymConst{\HOLTokenIn{}} \HOLFreeVar{B} \HOLSymConst{\HOLTokenConj{}} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{n}. \HOLBoundVar{n} \HOLSymConst{\HOLTokenIn{}} \HOLFreeVar{A} \HOLSymConst{\HOLTokenImp{}} \HOLSymConst{\HOLTokenNeg{}}\HOLFreeVar{R} \HOLBoundVar{n} \HOLBoundVar{k}
\end{alltt}

With all these results, finally we can prove the ``coarsest congruence
contained in $\approx$'' theorem for finite-state CCS:
\begin{alltt}
\HOLTokenTurnstile{} \HOLConst{finite_state} \HOLFreeVar{p} \HOLSymConst{\HOLTokenConj{}} \HOLConst{finite_state} \HOLFreeVar{q} \HOLSymConst{\HOLTokenImp{}}
   (\HOLFreeVar{p} \HOLSymConst{\HOLTokenObsCongr} \HOLFreeVar{q} \HOLSymConst{\HOLTokenEquiv{}} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{r}. \HOLFreeVar{p} \HOLSymConst{+} \HOLBoundVar{r} \HOLSymConst{\HOLTokenWeakEQ} \HOLFreeVar{q} \HOLSymConst{+} \HOLBoundVar{r})
\end{alltt}

For contraction and rooted contraction, the situation (and proof
steps) is exactly the same:
\begin{displaymath}
\xymatrix{
{\textrm{Contraction } (\succeq_{\mathrm{bis}})} & {\textrm{Sum
    contraction } (\succeq_{\mathrm{bis}}^+)} \ar@/^3ex/[ldd]^{\subseteq ?}\\
{\textrm{Contraction precongruence } ([\succeq_{\mathrm{bis}}])}
\ar[u]^{\subseteq} \ar[ru]^{\subseteq} \\
{\textrm{Rooted contraction } (\succeq_{\mathrm{bis}}^c)} \ar[u]^{\subseteq}
}
\end{displaymath}

And we got two versions of theorem saying rooted contraction is the
coarsest precongruence of (bisimilarity) contraction:
\begin{alltt}
\HOLTokenTurnstile{} \HOLConst{free_action} \HOLFreeVar{p} \HOLSymConst{\HOLTokenConj{}} \HOLConst{free_action} \HOLFreeVar{q} \HOLSymConst{\HOLTokenImp{}}
   (\HOLFreeVar{p} \HOLSymConst{\HOLTokenObsContracts} \HOLFreeVar{q} \HOLSymConst{\HOLTokenEquiv{}} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{r}. \HOLFreeVar{p} \HOLSymConst{+} \HOLBoundVar{r} \HOLSymConst{\HOLTokenContracts{}} \HOLFreeVar{q} \HOLSymConst{+} \HOLBoundVar{r})
\HOLTokenTurnstile{} \HOLConst{finite_state} \HOLFreeVar{p} \HOLSymConst{\HOLTokenConj{}} \HOLConst{finite_state} \HOLFreeVar{q} \HOLSymConst{\HOLTokenImp{}}
   (\HOLFreeVar{p} \HOLSymConst{\HOLTokenObsContracts} \HOLFreeVar{q} \HOLSymConst{\HOLTokenEquiv{}} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{r}. \HOLFreeVar{p} \HOLSymConst{+} \HOLBoundVar{r} \HOLSymConst{\HOLTokenContracts{}} \HOLFreeVar{q} \HOLSymConst{+} \HOLBoundVar{r})
\end{alltt}

\subsection{Semantic context, guardness and (pre)congruence}

We need to find a suitable formal definition of semantic
context. There're multiple ways. Here we have chosen to use $\lambda$-expressions (typed
$CCS\rightarrow CCS$) to represent a (multi-hole) semantic
context. The definition is inductive:
\begin{alltt}
\HOLConst{CONTEXT} (\HOLTokenLambda{}\HOLBoundVar{t}. \HOLBoundVar{t})
\HOLConst{CONTEXT} (\HOLTokenLambda{}\HOLBoundVar{t}. \HOLFreeVar{p})
\HOLConst{CONTEXT} \HOLFreeVar{e} \HOLSymConst{\HOLTokenImp{}} \HOLConst{CONTEXT} (\HOLTokenLambda{}\HOLBoundVar{t}. \HOLFreeVar{a}\HOLSymConst{..}\HOLFreeVar{e} \HOLBoundVar{t})
\HOLConst{CONTEXT} \HOLFreeVar{e\sb{\mathrm{1}}} \HOLSymConst{\HOLTokenConj{}} \HOLConst{CONTEXT} \HOLFreeVar{e\sb{\mathrm{2}}} \HOLSymConst{\HOLTokenImp{}} \HOLConst{CONTEXT} (\HOLTokenLambda{}\HOLBoundVar{t}. \HOLFreeVar{e\sb{\mathrm{1}}} \HOLBoundVar{t} \HOLSymConst{+} \HOLFreeVar{e\sb{\mathrm{2}}} \HOLBoundVar{t})
\HOLConst{CONTEXT} \HOLFreeVar{e\sb{\mathrm{1}}} \HOLSymConst{\HOLTokenConj{}} \HOLConst{CONTEXT} \HOLFreeVar{e\sb{\mathrm{2}}} \HOLSymConst{\HOLTokenImp{}} \HOLConst{CONTEXT} (\HOLTokenLambda{}\HOLBoundVar{t}. \HOLFreeVar{e\sb{\mathrm{1}}} \HOLBoundVar{t} \HOLSymConst{\ensuremath{\parallel}} \HOLFreeVar{e\sb{\mathrm{2}}} \HOLBoundVar{t})
\HOLConst{CONTEXT} \HOLFreeVar{e} \HOLSymConst{\HOLTokenImp{}} \HOLConst{CONTEXT} (\HOLTokenLambda{}\HOLBoundVar{t}. \HOLSymConst{\ensuremath{\nu}} \HOLFreeVar{L} (\HOLFreeVar{e} \HOLBoundVar{t}))
\HOLConst{CONTEXT} \HOLFreeVar{e} \HOLSymConst{\HOLTokenImp{}} \HOLConst{CONTEXT} (\HOLTokenLambda{}\HOLBoundVar{t}. \HOLConst{relab} (\HOLFreeVar{e} \HOLBoundVar{t}) \HOLFreeVar{rf})
\end{alltt}

Under above definition, we can formally define the concept of
``precongruence'' and ``congruence'' in the following ways:
\begin{alltt}
\HOLConst{precongruence} \HOLFreeVar{R} \HOLSymConst{\HOLTokenEquiv{}}
\HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{x} \HOLBoundVar{y} \HOLBoundVar{ctx}. \HOLConst{CONTEXT} \HOLBoundVar{ctx} \HOLSymConst{\HOLTokenImp{}} \HOLFreeVar{R} \HOLBoundVar{x} \HOLBoundVar{y} \HOLSymConst{\HOLTokenImp{}} \HOLFreeVar{R} (\HOLBoundVar{ctx} \HOLBoundVar{x}) (\HOLBoundVar{ctx} \HOLBoundVar{y})
\HOLConst{congruence} \HOLFreeVar{R} \HOLSymConst{\HOLTokenEquiv{}} \HOLConst{equivalence} \HOLFreeVar{R} \HOLSymConst{\HOLTokenConj{}} \HOLConst{precongruence} \HOLFreeVar{R}
\end{alltt}

A \emph{weak guarded} context is a context in which all holes appears
in the sub-expression of prefixed processes. We can easily define it
following the same idea:
\begin{alltt}
\HOLConst{WG} (\HOLTokenLambda{}\HOLBoundVar{t}. \HOLFreeVar{p})
\HOLConst{CONTEXT} \HOLFreeVar{e} \HOLSymConst{\HOLTokenImp{}} \HOLConst{WG} (\HOLTokenLambda{}\HOLBoundVar{t}. \HOLFreeVar{a}\HOLSymConst{..}\HOLFreeVar{e} \HOLBoundVar{t})
\HOLConst{WG} \HOLFreeVar{e\sb{\mathrm{1}}} \HOLSymConst{\HOLTokenConj{}} \HOLConst{WG} \HOLFreeVar{e\sb{\mathrm{2}}} \HOLSymConst{\HOLTokenImp{}} \HOLConst{WG} (\HOLTokenLambda{}\HOLBoundVar{t}. \HOLFreeVar{e\sb{\mathrm{1}}} \HOLBoundVar{t} \HOLSymConst{+} \HOLFreeVar{e\sb{\mathrm{2}}} \HOLBoundVar{t})
\HOLConst{WG} \HOLFreeVar{e\sb{\mathrm{1}}} \HOLSymConst{\HOLTokenConj{}} \HOLConst{WG} \HOLFreeVar{e\sb{\mathrm{2}}} \HOLSymConst{\HOLTokenImp{}} \HOLConst{WG} (\HOLTokenLambda{}\HOLBoundVar{t}. \HOLFreeVar{e\sb{\mathrm{1}}} \HOLBoundVar{t} \HOLSymConst{\ensuremath{\parallel}} \HOLFreeVar{e\sb{\mathrm{2}}} \HOLBoundVar{t})
\HOLConst{WG} \HOLFreeVar{e} \HOLSymConst{\HOLTokenImp{}} \HOLConst{WG} (\HOLTokenLambda{}\HOLBoundVar{t}. \HOLSymConst{\ensuremath{\nu}} \HOLFreeVar{L} (\HOLFreeVar{e} \HOLBoundVar{t}))
\HOLConst{WG} \HOLFreeVar{e} \HOLSymConst{\HOLTokenImp{}} \HOLConst{WG} (\HOLTokenLambda{}\HOLBoundVar{t}. \HOLConst{relab} (\HOLFreeVar{e} \HOLBoundVar{t}) \HOLFreeVar{rf})
\end{alltt}
(Notice the differences between a weak guarded context and a normal
one: $\lambda t. t$ is not weakly guarded as the variable is directly
exposed without any prefixed action. And $\lambda t. a.e[t]$ is weakly
guarded as long as $e[\cdot]$ is a semantic context, no necessary weakly guaded.)

It's a similar process to define strongly guarded context
(\texttt{SG}), sequential context (\texttt{SEQ}) and their
variants without direct sums (\texttt{GCONTEXT}, \texttt{WGS},
\texttt{GSEQ}). Some lemmas about their relationships have very long
(but trivial) formal proofs due to the combinatorial explosion of
inductions on their structures. For example, one such 
lemma says, for any semantic context $E$ which is both strongly guarded
and sequential (no direct sums) and another sequential context $H$ (no
direct sums), the composition $H \circ E$ is still both strongly
guarded and sequential (no direct sums):
\begin{alltt}
SG_GSEQ_combin:
\HOLTokenTurnstile{} \HOLConst{SG} \HOLFreeVar{E} \HOLSymConst{\HOLTokenConj{}} \HOLConst{GSEQ} \HOLFreeVar{E} \HOLSymConst{\HOLTokenImp{}} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{H}. \HOLConst{GSEQ} \HOLBoundVar{H} \HOLSymConst{\HOLTokenImp{}} \HOLConst{SG} (\HOLBoundVar{H} \HOLSymConst{\HOLTokenCompose} \HOLFreeVar{E}) \HOLSymConst{\HOLTokenConj{}} \HOLConst{GSEQ} (\HOLBoundVar{H} \HOLSymConst{\HOLTokenCompose} \HOLFreeVar{E})
\end{alltt}

\subsection{Milner's ``unique solution of equations'' theorems}

Once the representation issue of CCS equations is resolved, the actual
proofs of Milner's unique solution of equations theorems is not very
interesting from the view of theorem proving, although it's a 
precise proof engineering work producing quite long proofs.
Since we have chosen to use semantic context to represent
single-variable equations, an equation like $P \sim E\{P/X\}$ now
becomes $P \sim E[P]$, in which $E$ is a (multi-hole) semantic
context, and there's no need to say it ``contains at most the variable
$X$'' any more, as equation variable doesn't appear in $E$ at
all. Under these simplifications, the formal proof of Milner's unique
solution of equations theorem for $\sim$ is a literal mapping for informal
proofs based on bisimulation upto $\sim$, induction and case analysis
of weakly guarded contexts. Below is the formal version of Lemma 4.13
and Proposition 4.14 in Milner's book:

\begin{alltt}
STRONG_UNIQUE_SOLUTION:
\HOLTokenTurnstile{} \HOLConst{WG} \HOLFreeVar{E} \HOLSymConst{\HOLTokenImp{}} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{P} \HOLBoundVar{Q}. \HOLBoundVar{P} \HOLSymConst{\HOLTokenStrongEQ} \HOLFreeVar{E} \HOLBoundVar{P} \HOLSymConst{\HOLTokenConj{}} \HOLBoundVar{Q} \HOLSymConst{\HOLTokenStrongEQ} \HOLFreeVar{E} \HOLBoundVar{Q} \HOLSymConst{\HOLTokenImp{}} \HOLBoundVar{P} \HOLSymConst{\HOLTokenStrongEQ} \HOLBoundVar{Q}
\end{alltt}

\begin{alltt}
WEAK_UNIQUE_SOLUTION:
\HOLTokenTurnstile{} \HOLConst{SG} \HOLFreeVar{E} \HOLSymConst{\HOLTokenConj{}} \HOLConst{GSEQ} \HOLFreeVar{E} \HOLSymConst{\HOLTokenImp{}} \HOLSymConst{\HOLTokenForall{}}\HOLBoundVar{P} \HOLBoundVar{Q}. \HOLBoundVar{P} \HOLSymConst{\HOLTokenWeakEQ} \HOLFreeVar{E} \HOLBoundVar{P} \HOLSymConst{\HOLTokenConj{}} \HOLBoundVar{Q} \HOLSymConst{\HOLTokenWeakEQ} \HOLFreeVar{E} \HOLBoundVar{Q} \HOLSymConst{\HOLTokenImp{}} \HOLBoundVar{P} \HOLSymConst{\HOLTokenWeakEQ} \HOLBoundVar{Q}
\end{alltt}

>>>>

@@

In the second half of this paper, we present a formalization of CCS
including the new concepts and theorems proposed in the first half of
this paper. The main purpose is to convince the readers that, there's no flaw in the informal proofs. The
formalization work is based on a formalization of CCS in HOL theorem
prover (HOL4) \cite{slind2008brief}, described in second author's MSc thesis
\cite{Tian:2017wrba} and now lives in the official examples folder of HOL4 source
code\footnote{\url{https://github.com/HOL-Theorem-Prover/HOL}}.
The work is essentially a porting and extension of the early CCS
formalization by Monica Nesi \cite{Nesi} in HOL88.

Higher Order Logic (or ``HOL Logic'') \cite{hollogic} stands for simple-typed $\lambda$-calculus plus Hibert
choice operator, axiom of infinity, and rank-1 polymorphim (type
variables). HOL4 implemented this logic without any extension, while
some other theorem provers in HOL family (e.g. Isabelle/HOL) has
various extensions. The HOL Logic is considerability simpler than logic
foundations in most other theorem provers, e.g. Coq. As a result,
formal theories in HOL can be easily migrated (sometimes even
automatically) into other theorem provers.

@@

But such a simplification doesn't hurt the rigorousness of our entire
work, as once the single-equation version of the theorem is
proven, the general case is just a routine
adaptation for paper proofs.
It's not easy to formalize multi-variable CCS equations with
introducing a large number of definitions and intermediate
results. It's easy, however, to directly use the semantics context to
represent single-variable equations without the need of formalizing
the concept of ``equation'' at all.

There're two major limitations in this work. One is due to HOL, we only support Finitary CCS, i.e.
no infinite sum nor infinite parallel composition.
The other limitation is, equation variables (or holes in semantics contexts)
cannot appear inside recursion operator closure. However, these limitations do not
cause any difference in the proof of main results represented in this paper.
